Active Learning for SAT Solver Benchmarking {\large (extended and revised version)}

### 2025-02-14 paper_jar/paper.tex
Disable tracked changes
Minor language changes
For consistency or other reasons.

% Highlight changes for re-submission (might be removed later)
\usepackage{soul}
\soulregister\cite7 % some hacking to make highlighting work in special situations, e.g., if text contains references: https://tex.stackexchange.com/q/139463
\soulregister\ref7
\soulregister\pageref7
\newcommand{\showchanges}{1} % set to 0 or 1
\newcommand{\change}[1]{\ifnum \showchanges > 0 \sethlcolor{yellow}\hl{#1}\else#1\fi}

@@

The uncertainty sampling strategy selects the instance closest to the model's decision boundary.
This is achieved by selecting the instance $e \in \mathcal{I} \setminus \tilde{\mathcal{I}}$ that minimizes the certainty~$U(e)$, which is specified by the following equation:
%
\begin{equation*}
  \operatorname{U}(e) := \left\lvert \frac{1}{k} - \max_{n \in \left\lbrace 1, \dots, k \right\rbrace} f_{\hat{a}}'(e)_{n} \right\rvert \text{.}
\end{equation*}
\change{Intuitively, uncertainty sampling selects the instances where it is most unclear in which runtime cluster the new solver resides.
For example, if the new solver behaves similarly to two known solvers on the already run experiments, it is likely that this sampling strategy selects instances where these two solvers have runtimes that are not within the same cluster.}

@@

As a baseline, Table~\ref{tab:regression} shows the prediction performance of two regression models, i.e., a random forest and a multilayer perceptron.
We employ these models to estimate a new solver's raw runtime as the prediction target.
To this end, we use 46 instance features \change{(Section~\ref{sec:exdesign:data})} and 27 known solver runtimes as inputs for the predictions.
\change{For this experiment, the data is split into a training set (80\% of all data) on which the target runtime is known and a test set (20\% of all data) on which we predict the runtimes of the new solver.}
We repeat this evaluation for each solver as the prediction target and report the average and standard deviation of prediction performance over the 28 solvers \change{in Table~\ref{tab:regression}}.

@@

Table~\ref{tab:timeout-prediction} shows the prediction performance for estimating whether a solver times out or not.
This binary prediction target represents a very strong discretization of runtimes.
We compare a variety of prediction models that pursue different learning paradigms.
Each is trained with either instance features only (second column) or instance and runtime features (third column).
\change{For this experiment, the data is split into a training set (80\% of all data) on which the target label is known and a test set (20\% of all data) on which we predict the runtime label of the new solver.}
As the evaluation metric for prediction performance, we use Matthews Correlation Coefficient (MCC)~\cite{gorodkin2004comparing, matthews1975comparison}, which reaches its maximum of~1 for a perfect prediction, is~0 for random guessing, and has a minimum of~-1.

@@

Next, we use the runtimes labels from this discretization procedure as prediction targets.
Table~\ref{tab:cluster-prediction} shows the prediction performance for the best classification models from Table~\ref{tab:timeout-prediction} and $k \in \lbrace 3, 4, 5 \rbrace$ clusters.
We train the prediction models either with 46 instance features \change{(Section~\ref{sec:exdesign:data})} and 27 known solver runtimes (second column) or additionally including 27 features representing the known $C_k$ cluster labels of other solvers (third column).
\change{For this experiment, the data is split into a training set (80\% of all data) on which the target cluster label is known and a test set (20\% of all data) on which we predict the cluster label of the new solver.}

@@

In this section, we evaluate our active learning framework.
\change{While we split the data into a fixed train and test set to evaluate performance in Section~\ref{sec:exdesign:disc-pred}, in this section, the target runtime labels of the new solver are only known for the experiments that have already been selected by active-learning.}
First, we analyze and tune the different subroutines of our framework on the tuning dataset (cf.~Section~\ref{sec:eval:hyper}).
Next, we evaluate the best configurations on the full dataset (cf.~Section~\ref{sec:eval:full}).
Finally, we analyze the importance of instance families for our framework (cf.~Section~\ref{sec:eval:instance}).

@@

Both of our AL approaches outperform random sampling.
However, the performance differences are not significant for a Wilcoxon signed-rank test with $\alpha = 0.05$ and also depend on the fraction of the runtime sampled (cf.~Fig.~\ref{fig:annitraincolorselection}).
\change{However, an advantage of our approach is that it indicates when to stop adding new instances.
Recall that Figure~\ref{fig:annitrainoptgoalacc} shows the trade-offs of the amount of sampled instances and the ranking accuracy.
Active-learning helps to identify the optimal region as sampling too many instances only provides marginal improvements compared to its cost.}

@@

\change{For predictions within our framework, we use a database of 56 instance features\footnote{\url{https://udopia.github.io/gbdc/doc/extractors/CNFBaseFeatures.html}} from the Global Benchmark Database~(GBD)~\cite{IserJ24}.
Table~\ref{tab:features} briefly summarizes all 56 features.
The placeholder \texttt{<n>} can take values within the range 1 to 9 and the placeholder \texttt{<dist>} is one of \texttt{mean}, \texttt{variance}, \texttt{min}, \texttt{max}, or \texttt{entropy}.
Accounting for all possible instantiations of \texttt{<n>} and \texttt{<dist>}, there are 56 features in total.}

\begin{table}[h]
    \centering
    \caption{
      \change{List of all 56 instance features, where \texttt{<dist>} is one of \texttt{mean}, \texttt{variance}, \texttt{min}, \texttt{max}, or \texttt{entropy}.}
    }
    \label{tab:features}
    \begin{tabular}{
      >{\arraybackslash}p{0.22\textwidth}
      >{\arraybackslash}p{0.21\textwidth}
      >{\arraybackslash}p{0.46\textwidth}
    }
      \toprule
      Category & Feature Name & Description \\
      \midrule
      Basic statistics & \verb|clauses| & Number of clauses \\
      & \verb|variables| & Number of variables \\
      \midrule
      Clause types & \verb|cls<n>| & Number of clauses with \verb|<n>| literals for \verb|<n>| in 1--9 \\
      & \verb|cls10p| & Number of clauses with 10 or more literals \\
      & \verb|horn| & Number of Horn clauses \\
      & \verb|invhorn| & Number of inverted Horn clauses \\
      & \verb|positive| & Number of positive clauses \\
      & \verb|negative| & Number of negative clauses \\
      \midrule
      Horn proximity & \verb|hornvars_<dist>| & Distribution of variable occurrence in horn clauses \\
      & \verb|invhornvars_<dist>| & Distribution of variable occurrence in inverted horn clauses \\
      \midrule
      Polarity balance & \verb|balancecls_<dist>| & Distributions of fractions of number of positive/negative literals in clauses \\
      & \verb|balancevars_<dist>| & Distributions of fractions of number of positive/negative literals of variables \\
      \midrule
      Variable-clause graph & \verb|vcg_vdegree_<dist>| & Variable degree distribution \\
      & \verb|vcg_cdegree_<dist>| & Clause degree distribution \\
      \midrule
      Variable graph & \verb|vg_degree_<dist>| & Degree distribution \\
      \midrule
      Clause graph & \verb|cg_degree_<dist>| & Degree distribution \\
      \bottomrule
    \end{tabular}
\end{table}

>>>>

% Highlight changes for re-submission (might be removed later)
\usepackage{soul}
\soulregister\cite7 % some hacking to make highlighting work in special situations, e.g., if text contains references: https://tex.stackexchange.com/q/139463
\soulregister\ref7
\soulregister\pageref7
\newcommand{\showchanges}{0} % set to 0 or 1
\newcommand{\change}[1]{\ifnum \showchanges > 0 \sethlcolor{yellow}\hl{#1}\else#1\fi}

@@

The uncertainty sampling strategy selects the instance closest to the model's decision boundary.
This is achieved by selecting the instance $e \in \mathcal{I} \setminus \tilde{\mathcal{I}}$ that minimizes the certainty~$U(e)$, which is specified by the following equation:
%
\begin{equation*}
  \operatorname{U}(e) := \left\lvert \frac{1}{k} - \max_{n \in \left\lbrace 1, \dots, k \right\rbrace} f_{\hat{a}}'(e)_{n} \right\rvert \text{.}
\end{equation*}
\change{Intuitively, uncertainty sampling selects the instances where it is most unclear in which runtime cluster the new solver resides.
For example, if the new solver behaves similarly to two known solvers in the already run experiments, it is likely that this sampling strategy selects instances where these two solvers have runtimes that are not within the same cluster.}

@@

As a baseline, Table~\ref{tab:regression} shows the prediction performance of two regression models, i.e., a random forest and a multilayer perceptron.
We employ these models to estimate a new solver's raw runtime as the prediction target.
To this end, we use 46 instance features \change{(cf.~Section~\ref{sec:exdesign:data})} and 27 known solver runtimes as inputs for the predictions.
\change{For this experiment, we split the data into a training set (80\% of all data) on which the target runtime is known and a test set (20\% of all data) on which we predict the runtimes of the new solver.}
We repeat this evaluation for each solver as the prediction target and report the average and standard deviation of prediction performance over the 28 solvers \change{in Table~\ref{tab:regression}}.

@@

Table~\ref{tab:timeout-prediction} shows the prediction performance for estimating whether a solver times out or not.
This binary prediction target represents a very strong discretization of runtimes.
We compare a variety of prediction models that pursue different learning paradigms.
Each is trained with either instance features only (second column) or instance and runtime features (third column).
\change{For this experiment, we split the data into a training set (80\% of all data) on which the target label is known and a test set (20\% of all data) on which we predict the runtime label of the new solver.}
As the evaluation metric for prediction performance, we use Matthews Correlation Coefficient (MCC)~\cite{gorodkin2004comparing, matthews1975comparison}, which reaches its maximum of~1 for a perfect prediction, is~0 for random guessing, and has a minimum of~-1.

@@

Next, we use the runtimes labels from this discretization procedure as prediction targets.
Table~\ref{tab:cluster-prediction} shows the prediction performance for the best classification models from Table~\ref{tab:timeout-prediction} and $k \in \lbrace 3, 4, 5 \rbrace$ clusters.
We train the prediction models either with 46 instance features \change{(cf.~Section~\ref{sec:exdesign:data})} and 27 known solver runtimes (second column) or additionally including 27 features representing the known $C_k$ cluster labels of other solvers (third column).
\change{For this experiment, we split the data into a training set (80\% of all data) on which the target cluster label is known and a test set (20\% of all data) on which we predict the cluster label of the new solver.}

@@

In this section, we evaluate our active learning framework.
\change{While we split the data into a fixed train and test set to evaluate performance in Section~\ref{sec:exdesign:disc-pred}, in this section, the target runtime labels of the new solver are only known for the experiments that have already been selected by active learning.}
First, we analyze and tune the different subroutines of our framework on the tuning dataset (cf.~Section~\ref{sec:eval:hyper}).
Next, we evaluate the best configurations on the full dataset (cf.~Section~\ref{sec:eval:full}).
Finally, we analyze the importance of instance families for our framework (cf.~Section~\ref{sec:eval:instance}).

@@

Both of our AL approaches outperform random sampling.
However, the performance differences are not significant for a Wilcoxon signed-rank test with $\alpha = 0.05$ and also depend on the fraction of the runtime sampled (cf.~Fig.~\ref{fig:annitraincolorselection}).
\change{However, an advantage of our approach is that it indicates when to stop adding new instances.
Recall that Fig.~\ref{fig:annitrainoptgoalacc} shows the trade-off of the number of sampled instances and the ranking accuracy.
Active learning helps to identify the optimal region as sampling too many instances only provides marginal improvements compared to its cost.}

@@

\change{For predictions within our framework, we use a database of 56 instance features\footnote{\url{https://udopia.github.io/gbdc/doc/extractors/CNFBaseFeatures.html}} from the Global Benchmark Database~(GBD)~\cite{IserJ24}.
Table~\ref{tab:features} briefly summarizes all 56 features.
The placeholder \texttt{<n>} can take values within the range from 1 to 9, and the placeholder \texttt{<dist>} is one of \texttt{mean}, \texttt{variance}, \texttt{min}, \texttt{max}, or \texttt{entropy}.
Accounting for all possible instantiations of \texttt{<n>} and \texttt{<dist>}, there are 56 features in total.}

\begin{table}[h]
    \centering
    \caption{
      \change{List of all 56 instance features, where \texttt{<dist>} is one of \texttt{mean}, \texttt{variance}, \texttt{min}, \texttt{max}, or \texttt{entropy}.}
    }
    \label{tab:features}
    \begin{tabular}{
      >{\arraybackslash}p{0.22\textwidth}
      >{\arraybackslash}p{0.21\textwidth}
      >{\arraybackslash}p{0.46\textwidth}
    }
      \toprule
      Category & Feature Name & Description \\
      \midrule
      Basic statistics & \verb|clauses| & Number of clauses \\
      & \verb|variables| & Number of variables \\
      \midrule
      Clause types & \verb|cls<n>| & Number of clauses with \verb|<n>| literals for \verb|<n>| in 1--9 \\
      & \verb|cls10p| & Number of clauses with 10 or more literals \\
      & \verb|horn| & Number of Horn clauses \\
      & \verb|invhorn| & Number of inverted Horn clauses \\
      & \verb|positive| & Number of positive clauses \\
      & \verb|negative| & Number of negative clauses \\
      \midrule
      Horn proximity & \verb|hornvars_<dist>| & Distribution of variable occurrence in horn clauses \\
      & \verb|invhornvars_<dist>| & Distribution of variable occurrence in inverted Horn clauses \\
      \midrule
      Polarity balance & \verb|balancecls_<dist>| & Distributions of fractions of number of positive/negative literals in clauses \\
      & \verb|balancevars_<dist>| & Distributions of fractions of number of positive/negative literals of variables \\
      \midrule
      Variable-clause graph & \verb|vcg_vdegree_<dist>| & Variable degree distribution \\
      & \verb|vcg_cdegree_<dist>| & Clause degree distribution \\
      \midrule
      Variable graph & \verb|vg_degree_<dist>| & Degree distribution \\
      \midrule
      Clause graph & \verb|cg_degree_<dist>| & Degree distribution \\
      \bottomrule
    \end{tabular}
\end{table}
